A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-08-10 papers/water-SAFT/paper.tex
changes to paper and fixed a citation Tested everything on kipu.

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms---ideal gas, hard-sphere repulsion and
dispersion---encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy for fluids that (unlike water) are chain polymers.  While a
number of formulations of SAFT have been published, we will focus on
SAFT-VR\cite{gil-villegas-1997-SAFT-VR}, which was used by Clark
\emph{et al} to construct an optimal SAFT model for
water\cite{clark2006developing}. All but one of the six empirical
parameters used in the functional introduced in this paper are taken
directly from this Clark \emph{et al} paper.  As an example of the
power of this model, it predicts an enthalpy of vaporization at
100$^\circ$C of $\Delta H_{vap}$= 39.41~kJ/mol, compared with the
experimental value $\Delta H_{vap}$= 40.65~kJ/mol\cite{nistwater},
with an error of only a few percent.  We show a phase diagram for
this optimal SAFT model for water in
Figure~\ref{fig:pressure-with-isotherms}, which demonstrates that its
vapor pressure as a function of temperature is very accurate, while
the liquid density shows larger discrepancies.  The critical point is
very poorly described, which is a common failing of models that are
based on a mean-field expansion.

@@

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
We show in Figure~\ref{fig:energy-vs-diameter} the excess chemical
potential at room temperature, scaled by the solvent-accessible
surface area of the hard rod, plotted as a function of hard-rod
radius, which we define as the radius from which water is excluded.
For rods with radius larger than 0.5~nm or so, this reaches a maximum value of
75~mN/m, which is slightly higher than the bulk surface.  In the limit
of very large rods, this value will decrease and approach the bulk
surface value.  As seen in the inset of
Fig.~\ref{fig:energy-vs-diameter}, for rods with very small radius (less than
about 0.5~\AA) the excess chemical potential is proportional to volume,
as required by the contact-value theorem (see Equation~\ref{contactvaluethm}).

@@

We now look at the more interesting problem of two parallel hard rods
in water, separated by a distance $d$, as shown in
Figure~\ref{fig:density-rods}.  At small separations there is only vapor
between the rods, but as the rods are pulled apart, the vapor region
expands until a critical separation is reached at which point liquid
water fills the region between the rods.
Figure~\ref{fig:density-rods} shows density profiles before and after
this transition for rods of radius 0.6~nm. This critical separation
for the transition to liquid depends on the radii of the rods, and is
about 0.65~nm for the rods shown in Figure~\ref{fig:density-rods}.
The critical separation will be different for a system where there is
attraction between the
rods and water.  At small separations, the water around the two
rods looks as if the two rods were one solid ``stadium''-shaped object
(a rectangle with semi-circles on both ends).

@@

Walther \emph{et al.}\cite{walther2004hydrodynamic} studied the
interactions between two carbon nanotubes, which are geometrically
similar to our hydrophobic rods.  They use molecular dynamics with
Lennard-Jones potentials for nanotubes of diameter 1.25~nm and
separations ranging from about 0.3~nm to 1.5~nm, and find that a
drying transition occurs, and find it to be at separations near
0.9~-~1.0~nm\cite{walther2004hydrodynamic}. This is larger than what
we find for hydrophobic rods of a similar size, as expected due to the
carbon-water interactions.  \fixme{Can you (Jess or Eric) double-check
  what the excluded-volume diameter is for them? The discrepancy we
  describe in the text is actually the opposite of what I would
  expect, but looking at the plot, I see something that is the
  opposite of what is described in the text.  We should really figure
  this out, and probably spend another phrase or two clarifying why it
  is the direction we expect.}

>>>>

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms---ideal gas, hard-sphere repulsion and
dispersion---encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy for fluids that are chain polymers.  While a
number of formulations of SAFT have been published, we will focus on
SAFT-VR\cite{gil-villegas-1997-SAFT-VR}, which was used by Clark
\emph{et al} to construct an optimal SAFT model for
water\cite{clark2006developing}. All but one of the six empirical
parameters used in the functional introduced in this paper are taken
directly from this Clark \emph{et al} paper.  As an example of the
power of this model, it predicts an enthalpy of vaporization at
100$^\circ$C of $\Delta H_{vap}$= 39.41~kJ/mol, compared with the
experimental value $\Delta H_{vap}$= 40.65~kJ/mol\cite{nistwater},
with an error of only a few percent.  We show a phase diagram for
this optimal SAFT model for water in
Figure~\ref{fig:pressure-with-isotherms}, which demonstrates that its
vapor pressure as a function of temperature is very accurate, while
the liquid density shows larger discrepancies.  The critical point is
very poorly described, which is a common failing of models that are
based on a mean-field expansion.

@@

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
In Figure~\ref{fig:energy-vs-diameter} we show the excess chemical
potential at room temperature, scaled by the solvent-accessible
surface area of the hard rod, plotted as a function of hard-rod
radius.  We define the hard-rods radius as the radius from which water
is excluded.
For rods with radius larger than 0.5~nm or so, this reaches a maximum value of
75~mN/m, which is slightly higher than the bulk surface.  In the limit
of very large rods, this value will decrease and approach the bulk
surface value.  As seen in the inset of
Fig.~\ref{fig:energy-vs-diameter}, for rods with very small radius (less than
about 0.5~\AA) the excess chemical potential is proportional to volume,
as required by the contact-value theorem (see Equation~\ref{contactvaluethm}).

@@

We now look at the more interesting problem of two parallel hard rods
in water, separated by a distance $d$, as shown in
Figure~\ref{fig:density-rods}.  At small separations there is only vapor
between the rods, but as the rods are pulled apart, the vapor region
expands until a critical separation is reached at which point liquid
water fills the region between the rods.
Figure~\ref{fig:density-rods} shows density profiles before and after
this transition for rods of radius 0.6~nm. This critical separation
for the transition to liquid depends on the radii of the rods, and is
about 0.65~nm for the rods shown in Figure~\ref{fig:density-rods}.
The critical separation will be different for a system where there is
attraction between the
rods and water.  At small separations, the shape of the water around the two
rods makes them appear as one solid ``stadium''-shaped object
(a rectangle with semi-circles on both ends).

@@

Walther \emph{et al.}\cite{walther2004hydrophobic} studied the
interactions between two carbon nanotubes, which are geometrically
similar to our hydrophobic rods, by using molecular dynamics with
Lennard-Jones potentials for nanotubes of diameter 1.25~nm and
separations ranging from about 0.3~nm to 1.5~nm. They compute the
force to constrain the nanotubes at different sepearation distances
and find that a drying transition begins at separations near 0.9~-~1.0~nm where two
layers of water molecules can exist in a metastable state between the
rods. This reduces to one layer at a spacing of 0.55~-~0.58~nm, and at
0.35~nm water is complete expelled and is in
equilibrium\cite{walther2004hydrophobic}. The final drying state is
smaller than what we find for hydrophobic rods of a similar size, as
expected due to the carbon-water interaction and the
van der Waals interaction between the rods that causes more attraction
than our rods.  Our model does not predict any of the transitional
states seen here\fixme{why?}.
